% Contributors: Suman Mulumudi, Jake Lee

\section{Background on Autoencoders and A Few Flavors}

Autoencoders are neural networks which perform dimensionality
reduction in an unsupervised manner by compressing their inputs into
a (generally lower-dimensional) latent space while mandating that the
original input can be reconstructed. This turns out to be an
extremely powerful architecture, enabling both the discovery of
well-structured latent spaces and for the separation of signal from
noise.

Formally, suppose an input space $X$, an encoder $f(x ; \theta_f): X
\to H$, and a decoder $g(h ; \theta_g): H \to X$, where $H$ is a
latent space. Further, suppose a regressive reconstruction loss $l(x,
x'): X \times X \to R$. The autoencoder is then optimized as:

\begin{align*}
    \theta_f^*, \theta_g^* = \argmin_{\theta_f, \theta_g} \sum_i l(x_i, g(f(x_i)))
\end{align*}

That is, the encoder $f$ transforms $x$ into a latent representation
$h$, which is then transformed by $g$ back into the original domain.
Generally, $H$ will be chosen to be of lower dimension than $X$ in
order to facilitate dimensionality reduction, and the use of a
reconstruction loss forces a representation in $H$ to retain
information about the original input. Encoder and decoder functions
can be as simple as linear layers, or as complex as deep neural
networks with non-linear activations in order to learn richer or more
abstract representations. Optimization of autoencoders is generally
done by stochastic gradient descent.

To better understand the process of autoencoding, it may be helpful
to view $f(x)$ as projecting points onto a manifold $H$, and $g(x)$
as re-projecting from the manifold back into the original space; with
this conceptualization, training an autoencoder is the process of
finding a low-dimensional manifold which can best represent the data
the data in $X$. Indeed, when the encoder and decoder are linear, the
optimal autoencoder recovers the PCA subspace, which we know to be
the optimal linear manifold for dimensionality reduction
\cite{alain2014regularized}.


In de-noising applications, the full auto-encoder $g(f(x))$ is used
for downstream tasks. In generative applications, the decoder $g(h)$
is used in downstream tasks. In dimensionality reduction and
re-representation applications, the latent space representation
$h=f(x)$ is useful for downstream tasks. These are discussed in more
detail below.


\subsection{Regularized Autoencoders}

Regularized autoencoders utilize regularization on the latent space,
or by introducing noise into the input, to better constrain the
geometry of the latent space.

One common regularized autoencoder is the Contracted Autoencoder
(CAE). CAEs modify the autoencoder loss to:

\begin{align*}
    L_{CAE}(x, x') = \sum_i \left ( l(x_i, \hat{x}_i) + \lambda \left | \frac{\partial f(x)}{\partial x} \right |^2_F \right )
\end{align*}

This additional term to regularize the encoder adds robustness to
perturbations of the input space. In general, an autoencoder must
learn representations which allow for differentiation of different
datapoints and allow for reconstruction, so a strong contractive loss
term (i.e. large $\lambda$) provides robustness by encouraging the
autoencoder to be robust to directions in the input space which are
not descriptive of the data (i.e. orthogonal to the manifold on which
the data lie) \cite{alain2014regularized}.


Another common autoencoder is the denoising autoencoder (DAE). Unlike
other autoencoders, DAEs are used to separate data from noise by
introducing stochastic noise into the input and regressing the
uncorrupted output with the goal of producing an autoencoder which
can denoise an input. That is, if $N(x)$ is a function which
introduces stochastic noise, then a DAE has the form:

\begin{align*}
    \theta_f^*, \theta_g^* = \argmin_{\theta_f, \theta_g} \sum_i l(x_i, g(f(N(x_i))))
\end{align*}

Indeed, the DAE can actually be seen as a form of a CAE, as it
encourages robustness to small perturbations of the input, but
applies regularization over the full encoder-decoder chain rather
than just the encoder \cite{alain2014regularized}.



\subsection{Variational Autoencoders}

Variational autoencoders (VAEs) are a generative variant of
traditional autoencoders that is regularized in to create a latent
space which can be easily sampled to generate new examples similar to
the original data. The primary intuition behind VAEs is to try to
model the dataset as being sampled from a well structured latent
distribution, generally a latent normal distribution $h \in H =
N(0,I)$. Using this distribution as a prior, the VAE encoder then
maps each datapoint $x \in X$ to a posterior distribution
$N(\vec{\mu}, \Sigma)$ \cite{doersch2016tutorial}.

Formally, we are aiming to maximize the probability of the data
$P(X)$ over a latent space $h \in H$:

\begin{align*}
    P(X) = \int P(X \mid h) P(h) dh
\end{align*}

$P(h)$ is a model of the prior (which we will constrain to $N(0,I)$),
and $P(X \mid h)$ models the posterior (which we will also constrain
to a normal distribution), and the VAE attempts to optimize the
posterior such that probability of all the data in the dataset is
maximized. Optimizing this directly is generally computationally
intractable because of the size of the latent space, but by observing
that most latent variables $h$ will not produce a given datapoint $x$
(i.e. each datapoint is mapped to a small region of the latent
space), we can re-formulate the problem such that we sample a more
limited region of the latent space. Namely, re-write the optimization
in terms of the KL-divergence $D_{KL}$ and introducing an additional
probability distribution, $Q$:

\begin{align*}
    \log P(X) - D_{KL}[Q(h \mid X) \mid P(h \mid X)] \\
    = E_{h \sim Q} [\log P(X \mid h)] - D_{KL}[Q(h \mid X) \mid P(h)]
\end{align*}

(A formal derivation of this can be found in
\cite{doersch2016tutorial} – the derivation is relatively simple, but
is not in itself illustrative for intuition and is therefore omitted
here.) In this formulation, the distribution $Q(h \mid X)$ is modeled
by the encoder, and the distribution $P(X \mid h)$ is modeled by the
decoder. Additionally, re-writing the optimization in terms of the
KL-Divergence displays an important intuition: the KL-Divergence is a
similarity measure between probability distributions, so any
minimization over a KL-Divergence forces the two probability
distributions to be similar.

Intuitively, $E_{h \sim Q} [\log P(X \mid h)]$ maximizes the
probability of the data $X$ given $h \in Q$, as required by the
original optimization problem. $D_{KL}[Q(h \mid X) \mid P(h)]$ forces
the encoded latent space distribution for the data to be similar to
the prior distribution, making it a reasonable distribution over
which to sample. We also note that the $D_{KL}[Q(h \mid X) \mid P(h
\mid X)]$ vanishes when $Q$ is sufficiently high-capacity (i.e. the
encoder and decoder distributions will naturally tend to match during
training), leaving a maximization over $\log P(X)$ as desired.

VAEs achieve model this by making two changes to the standard
autoencoder: encoding each datapoint to a multivariate Gaussian
distribution rather than a single point (i.e. the posterior), and
applying a KL-divergence loss to the resulting distribution (i.e.
enforcing the prior). Formally, the encoder is modified to output a
$\vec{\mu}_i$ and $\Sigma_i$ of a distribution, and during training
the decoder is fed a vector from the latent space sampled from the
distribution $N(\vec{\mu}_i, \Sigma_i)$. That is, the encoder and
decoder are re-formulated as:

\begin{align*}
    &f(x_i ;\theta_f) = \vec{\mu}_i, \Sigma_i \\
    &\gamma \in N(0,I) \\
    &g(\vec{\mu}_i + \gamma \Sigma_i; \theta_g) = \hat{x}_i \\
\end{align*}

And the loss is reformulated as:

\begin{align*}
    L_{VAE}(x_i, \hat{x}) = l(x_i, \hat{x}_i) - D_{KL}[N(\vec{\mu}, \Sigma) \mid N(0,I)]
\end{align*}

Where $l$ is again the reconstruction loss. Intuitively, the
KL-divergence enforces the prior distribution and prevents each
datapoint from collapsing to a single point representation in the
latent space, and models $D_{KL}[Q(h \mid X) \mid P(h)]$. Further,
computing the reconstruction loss over a single point sampled from
the predicted distribution $N(\vec{\mu}_i, \Sigma_i)$ serves as a way
to tractable approximate $E_{h \sim Q} [\log P(X \mid h)]$ and over
which one can apply stochastic gradient descent. (Note: $\Sigma$ is
also often constrained to be diagonal to ease computation of the
KL-Divergence.)

Once such an autoencoder is trained, the latent space can be sampled
to generate new outputs from the original distribution by producing
vectors $h \in N(0,I)$ and feeding these values $h$ to the decoder
$g(h)$.%Jake

\subsection{Conditional VAE}

While a VAE could generate outputs from the overall original
distribution by feeding latent space samples to the decoder, in
certain cases it may become necessary to be more specific in this
data generation process. For example, if trained on a dataset of
digit images, we may want to generate images of a specific digit
instead of any digit from the overall distribution. To accomplish
this, we turn to the Conditional VAE.

Whereas a standard VAE encoder and decoder only conditions on the
input space and latent space respectively, a Conditional VAE, also
conditions both on an additional variable $c$. That is, the encoder
and decoder are re-formulated as:

\begin{align*}
    &f(x_i, c ;\theta_f) = \vec{\mu}_i, \Sigma_i \\
    &\gamma \in N(0,I) \\
    &g(\vec{\mu}_i + \gamma \Sigma_i, c; \theta_g) = \hat{x}_i \\
\end{align*}

And the optimization becomes similarly conditioned on $c$. This
conditioning is simply accomplished by concatenating $c$ to the input
vector and latent vector for the encoder and decoder, respectively.
Now that the latent variable is distributed conditioned on $c$, we
can now feed the decoder a randomly sampled latent space vector and
specify a concatenated $c$ to generate from the specified
distribution.

\subsection{$\beta$-VAE}

While a VAE will learn a latent representation ideal for decoding,
there is no guarantee on the \textit{disentanglement} of the learned
features. If each variable in a latent representation is sensitive to
changes in a single generative factor, while being relatively
invariant to changes in other factors, then it is referred to as
\textit{disentangled}. We introduce the Beta VAE, which encourages
the learning of such representations \cite{higgins2017beta}. This is
particularly useful for interpretability of the latent space, as well
as fine-grained manipulation of generated outputs. For example, a
Beta VAE trained on images of faces may be able to modify a single
factor such as hair color or mouth shape in a generated image by
modifying a single value in the latent space representation.

To achieve this during training, a single hyperparameter $\beta$ is
introduced. When $\beta = 1$, the Beta VAE behaves equally to a
standard VAE. When $\beta > 1$ however, the capacity of the learning
channel is limited, and the network is forced to learn a more
efficient latent representation that is hopefully more disentangled.
This optimization is defined as:

\begin{align*}
    L_{\beta \text{-VAE}}(x_i, \hat{x}_i) = l(x_i, \hat{x}_i) - \beta D_{KL}[N(\vec{\mu}_i, \Sigma_i) \mid N(0,I)]
\end{align*}

This simple addition of the $\beta$ coefficient leads to a more
disentangled latent representation, but may come with a trade-off
with reconstruction fidelity. Therefore, a $\beta$ value must be
found with the best balance between the disentanglement of the latent
space and the quality of decoding.